---
title: "Intermediate statistics: introduction"
author: "Frank Edwards"
institute: School of Criminal Justice, Rutgers - Newark
date: "1/16/2024"
output: binb::metropolis
---

```{r message = FALSE, warning = FALSE, echo=FALSE}
library(tidyverse)
### configure for variable text size with chunk option
def.chunk.hook  <- knitr::knit_hooks$get("chunk")
knitr::knit_hooks$set(chunk = function(x, options) {
  x <- def.chunk.hook(x, options)
  ifelse(options$size != "normalsize", paste0("\\", options$size,"\n\n", x, "\n\n \\normalsize"), x)
})
knitr::opts_chunk$set(warning=FALSE, message=FALSE, tidy = TRUE, size = "small")
```


# Introductions: What are you planning to do with statistical models?

## Before we begin

Remember: All models are wrong, some are useful.

## What we will cover

- How to explore, visualize, and model diverse kinds of data with an emphasis on generalized linear models \pause
- How to program in R \pause
- Developing a workflow for producing replicable quantitative social science \pause
- Some advanced topics that are relevant for the kinds of data we're dealing with in the course

## My general approach to data analysis

1. Explore and visualize data \pause
2. Fit models \pause
3. Assess model fit \pause
4. Interpret and describe results through simulation

## Review the syllabus

https://f-edwards.github.io/intermediate_stats/

## Set up and workflow basics

1. Install R

cran.r-project.org

2. Install RStudio

posit.co/download/rstudio-desktop/

3. File structure and project organization basics

4. R console basics

# Break

# The Generalized Linear Model

## The linear model

We know we can model data as:

<!-- THIS IS THE LINEAR REGRESSION FORMULA -->
<!-- where epsilon is the error term -->

$$y = \beta_0 + \beta_1x_1 \cdots \beta_nx_n + \varepsilon$$ \pause

Or, more succinctly:

$$ y =  \mathbf{X} \mathbf{\beta} + \varepsilon $$ \pause

Where the likelihood for the outcome conditional on the data takes the form:

$$ Y|X \sim Normal(\mu, \sigma^2) $$ 

## Generalzing the linear model

The linear model: 

$$ Y|X \sim Normal(\mu, \sigma^2) $$

Can be written as a more general formulation for a likelihood function $f$

$$ Y|X \sim f(\mu, \sigma^2) $$ \pause

Now we can extend the (very) useful linear model to data with discrete outcomes

## Generalizing the linear model

An expected value $E(Y|X) = \mu$

A linear predictor: 

$$ \mathbf{X} \mathbf{\beta} $$ \pause

A link function $g$

$$ g(\mu) =  \mathbf{X} \mathbf{\beta} $$ \pause

$$ \mu =  g^{-1}(\mathbf{X} \mathbf{\beta}) $$

## From OLS to GLM

OLS:

$$ Y|X \sim Normal(\mu, \sigma^2) $$ 

GLM, for a likelihood function $f$ with parameters $\theta$: 

$$ Y|X \sim f(\theta) $$

## Models we'll consider this semester

- Binary data: logistic models \pause
- Categorical data: Multinomial models \pause
- Count data: Poisson and negative binomial models 

# Returning to the linear model

## What do we know about the linear regression model?

$$ y =  \mathbf{X} \mathbf{\beta} + \varepsilon $$
$$\varepsilon \sim Normal(0, \sigma^2)$$

OR

$$\mu =  \mathbf{X} \mathbf{\beta}$$
$$y \sim Normal(\mu, \sigma^2)$$

# Let's analyze some data

## Two ways to access course data

- All data is accessible through the the course website (see the data link, or data folder on the GitHub page)

## Read in 
```{r size = "tiny"}
library(tidyverse)
### data available in intermediate_stats/data/revenue_dat.csv
cj_budgets<-read_csv("http://tinyurl.com/revenuedata1")
```

## About the data

It documents police involved deaths, demographics, and local government budgets at the county-level for two time periods, 2007-11 and 2012-16. Sources include Fatal Encounters, American Community Survey 5-year data, Annual Survey of State and Local Government Finance, and Uniform Crime Reports.

## Evaluate the structure of the data

```{r size = "tiny"}
head(cj_budgets)
```

## Descriptives

```{r}
summary(cj_budgets)
```

## Visualize the distribution of deep poverty across counties with ggplot

```{r fig.width=6, fig.height=3}
ggplot(cj_budgets,
       aes(x = pop_pct_pov)) + 
  geom_histogram()
```

## Create a new variable using mutate()

```{r}
cj_budgets <- cj_budgets %>% 
  mutate(pop_prop_blk = pop_blk / pop_tot)
```

## Visualize a bivariate relationship with ggplot

```{r fig.width=6, fig.height=3, size = "tiny"}
ggplot(cj_budgets,
       aes(y = pop_pct_pov,
           x = pop_prop_blk)) +
  geom_point()
```

## Fitting a linear model with lm()

```{r}
model_1<-lm(pop_pct_pov ~ 
              pop_prop_blk,
            data =cj_budgets)
```

## Display the model fit

```{r size = "tiny"}
summary(model_1)
```

## Display the model fit (nicer)

```{r size = "tiny"}
library(broom)
tidy(model_1)
```

## Visualize the model fit

```{r fig.width=6, fig.height=2.5, size = "tiny"}
ggplot(cj_budgets,
       aes(y=pop_pct_pov, x=pop_prop_blk))+
  geom_point() + 
  geom_smooth(method = "lm", 
              formula = y~x) 
```

# Other relationships we could evaluate?

# HW 1 guidelines

## HW tips

- Homework assignments are at github.com/f-edwards/intermediate_stats/hw/
- 
- Work together!
- Google it: StackOverflow will become your best friend
- Accept that this is hard and you will probably struggle with it