---
title: "Multilevel models, part 2"
author: "Frank Edwards"
output: binb::metropolis
---

```{r setup, include=FALSE}
library(gridExtra)
library(tidyverse)
library(lme4)

set.seed(1)
select<-dplyr::select

knitr::opts_chunk$set(tidy = FALSE)
theme_set(theme_bw())
def.chunk.hook  <- knitr::knit_hooks$get("chunk")
knitr::knit_hooks$set(chunk = function(x, options) {
  x <- def.chunk.hook(x, options)
  ifelse(options$size != "normalsize", paste0("\\", options$size,"\n\n", x, "\n\n \\normalsize"), x)
})
knitr::opts_chunk$set(warning=FALSE, message=FALSE, tidy = F, size = "small")
```

## Arrests and poverty

Data for today (available at https://github.com/f-edwards/intermediate_stats/tree/master/data)

```{r}
dat<-read_csv("./data/violent_arrest_data.csv")
```

## The Distribution of poverty and crime across counties

```{r, echo = F}
ggplot(dat,
       aes(x = pop_pct_pov, y = violent_arrest_per1000)) + 
  geom_point()
```

## A simple model

This model, with no included parameters for group differences, is called a *complete pooling* model. Data from all groups is used to estimate a single intercept and single slope.

Where violent crime rates are *y*, poverty rates are *x*, counties are *c*, states are *s*

\[y_c = \beta_0 + \beta_1x_c + \varepsilon_{c}\]
\[\varepsilon \sim N(0, \sigma^2)\]

## Estimating the complete-pooling model

```{r}
m_complete_pooling <- lm(violent_arrest_per1000 ~ pop_pct_pov,
                         data = dat)
broom::tidy(m_complete_pooling)
```

## Visualizing the model

```{r size = "tiny", echo = F, fig.height = 4}
ggplot(dat,
       aes(x = pop_pct_pov,
           y = violent_arrest_per1000)) + 
  geom_point() + 
  geom_abline(aes(intercept = coef(m_complete_pooling)[1],
                  slope = coef(m_complete_pooling)[2]),
              color = "blue",
              linewidth = 2)
```

## The state fixed effects model

Let's estimate a model where NO information across states is included in our estimation of intercepts. 

- Every state gets their own intercept (computed as the exact within-state violent crime rate mean).

## Estimating the state fixed effects model

```{r, size = "tiny"}
m_state_fe<-lm(violent_arrest_per1000 ~ 
                 pop_pct_pov +
                 factor(stusps),
               data = dat)
broom::tidy(m_state_fe)
```

## What an intercept means in practice

Intercepts are computed as the average violent arrest rate in each state, when poverty is fixed at zero
- State intercepts adjust for differences in average violent arrests rates at the county-level *across states*

## Visualizing the fixed effects (state intercepts) model

```{r, echo = F}
m_state_fe<-lm(violent_arrest_per1000 ~ -1 +
                 pop_pct_pov +
                 factor(stusps),
               data = dat)

fake_dat<-expand_grid(stusps = unique(dat$stusps),
                      pop_pct_pov = 0:50)

fake_dat<- fake_dat %>%
  mutate(yhat = predict(m_state_fe, newdata = fake_dat))

ggplot(dat,
       aes(x = pop_pct_pov,
           y = violent_arrest_per1000)) +
  geom_point() +
  geom_line(data = fake_dat,
            aes(x = pop_pct_pov, y = yhat, group = stusps),
            alpha = 0.25, color = "blue")
```

## Another visualization

```{r echo = F}
ggplot(dat,
       aes(x = pop_pct_pov,
           y = violent_arrest_per1000)) +
  geom_point() +
  geom_line(data = fake_dat,
            aes(x = pop_pct_pov, y = yhat, group = stusps), 
            color = "blue") + 
  facet_wrap(~stusps)
```

## Do we think that the relationship between poverty and arrest rates is identical across states?

- Intercepts capture *average* arrest levels within the state. They adjust for variation in arrest rates for both high and low poverty counties

- Slopes capture *differences* in arrest rates between high and low poverty counties both within and across states. 

- What processes could account for variation in *intercepts* and/or *slopes* in our study?

# Varying slopes 

## Returning to the fixed effects (intercepts) model

Where *y* is a violent arrest rate, *x* is a poverty rate, *c* indicates a county, and *s* indicates a state

\[y_c = \beta{_0}_s + \beta_1 x_c + \varepsilon_c \]

What does this model do?

## Modifying the model to allow for varying slopes

Where *y* is a violent arrest rate, *x* is a poverty rate, *c* indicates a county, and *s* indicates a state

*Fixed effects intercepts*

\[y_c = \beta{_0}_s + \beta_1 x_c + \varepsilon_c \]

*Fixed effects intercepts AND slopes*

\[y_c = \beta{_0}_s + \beta{_1}_s x_c + \varepsilon_c \]

## Breaking down how the fixed effects interaction model works

We are now effectively adding an *interaction* for the poverty variable and the state variable. 

We can rewrite the regression model like this, where $\beta_s$ is the adjustment we add to each state's slope from the reference category ($\beta_2$).

\[y_c = (\beta_0 + \beta{_1}_s) + (\beta_2 + \beta{_3}_s) x_c + \varepsilon_c\]

Now our relationship between poverty ($x$) and arrest rates $y$ depends both on poverty levels AND on the state a county is in. That's an interaction.

\pause

We've redefined our slope parameter to $\beta_2 + \beta{_3}_s$

## Estimating the fixed effects interaction model in R

```{r size = "tiny"}
m_state_fe_int<-lm(violent_arrest_per1000 ~ 
                 pop_pct_pov *
                 factor(stusps),
               data = dat)
broom::tidy(m_state_fe_int)
```

How many betas do we have (so many betas...)????

## What the model is doing

```{r echo = F}
fake_data<-expand_grid(stusps = unique(dat$stusps),
                       pop_pct_pov = 0:50)

fake_data<-fake_data %>% 
  mutate(violent_arrest_per1000_hat = predict(m_state_fe_int, newdata = fake_data))

ggplot(fake_data,
       aes(x = pop_pct_pov,
           y = violent_arrest_per1000_hat,
           group = stusps)) + 
  geom_line(alpha = 0.25, color = "blue")
```

## What the model is doing

```{r echo = F}
ggplot(fake_data,
       aes(x = pop_pct_pov,
           y = violent_arrest_per1000_hat,
           group = stusps)) + 
  geom_point(data = dat, 
             aes(x = pop_pct_pov, y = violent_arrest_per1000),
             alpha = 0.25) + 
  geom_line(color = "blue") + 
  facet_wrap(~stusps) + 
  ylim(c(0, 60))
```

## Three approaches to pooling data

- Under a complete-pooling model, we estimate a single intercept *and slope* for the full data based on the average violent crime rate *across all counties*
- Under a no-pooling model (the fixed effects interaction model), we estimate an intercept AND slope for every state based *only* on the data from that state
- Under a partial-pooling model, we estimate an intercept AND slope for every state based based on a model that *assumes that state intercepts AND slopes are generated from an underlying probability distribution*

## A basic multilevel model (intercepts only)

Where violent crime rates $y$ are a function of poverty rates $x$ in state $s$, and county $c$

\[y_c = (\beta_0 + \alpha_s) + \beta_1x_{c} + \varepsilon_c\]
\[\varepsilon \sim N(0, \sigma^2)\]
\[\alpha \sim N(0, \sigma^2_\alpha)\]

## A A basic multilevel model (intercepts and slopes)

\[y_c = (\beta_0 + \alpha_s) + (\beta_1 + \delta_s)x_c + \varepsilon_c\]
\[\varepsilon \sim N(0, \sigma^2)\]
\[\alpha \sim N(0, \sigma^2_\alpha)\]
\[\delta \sim N(0, \sigma^2_\delta)\]

## Estimating the random intercepts and slopes model

```{r size = "tiny"}
library(lme4)
# RE specification gets complex
# google lmer cheat sheet for support
m_state_ml<-lmer(violent_arrest_per1000 ~ 
                   pop_pct_pov +
                   (1|stusps) + #intercepts for each state
                   (0 + pop_pct_pov|stusps), #slopes for each state, uncorrelated
                 data = dat) 
```

## Model output

```{r size = "tiny"}
summary(m_state_ml)
```

## What the model does

```{r echo = F}
fake_data<-expand_grid(stusps = unique(dat$stusps),
                       pop_pct_pov = 0:50)

fake_data<-fake_data %>% 
  mutate(violent_arrest_per1000_hat = 
           predict(m_state_ml, newdata = fake_data))

ggplot(fake_data,
       aes(x = pop_pct_pov,
           y = violent_arrest_per1000_hat,
           group = stusps)) + 
  geom_line(alpha = 0.5)
```

## What the model does

```{r echo = F}
ggplot(fake_data,
       aes(x = pop_pct_pov,
           y = violent_arrest_per1000_hat,
           group = stusps)) + 
  geom_line(alpha = 0.5) + 
  geom_point(data = dat, 
             aes(x = pop_pct_pov, y = violent_arrest_per1000),
             alpha = 0.25) + 
  geom_line(color = "blue") + 
  facet_wrap(~stusps) + 
  ylim(c(0, 60))
```

## Let's compare multilevel and fixed effects model fits

```{r message = F, size = "tiny"}
fake_data<-expand_grid(stusps = unique(dat$stusps),
                       pop_pct_pov = 0:50)
# generate yhat for both fe and ml models
fake_data<-fake_data %>% 
  mutate(fe = predict(m_state_fe_int, newdata = fake_data),
         ml = predict(m_state_ml, newdata = fake_data))
# format for plotting
fake_data<- fake_data %>% 
  pivot_longer(cols = fe:ml, 
               names_to = "model_type",
               values_to = "yhat")

head(fake_data)
```

## Let's compare multilevel and fixed effects model fits

```{r echo = F}
ggplot(fake_data, 
       aes(x = pop_pct_pov, y = yhat, group = stusps)) + 
  geom_line() + 
  facet_wrap(~model_type)
```

## Let's compare multilevel and fixed effects model fits

```{r echo = F}
ggplot(fake_data, 
       aes(x = pop_pct_pov, y = yhat, 
           color = model_type)) + 
  geom_line() + 
  geom_point(data = dat, 
             aes(x = pop_pct_pov, y = violent_arrest_per1000),
             color = "black",
             alpha = 0.25) +
  facet_wrap(~stusps) + 
  ylim(c(0, 50))
```

## Benefits of the multilevel approach: regularization

Our model *regularizes* (or shrinks) intercept and slope estimates by nudging them toward the global mean

This shift is larger when:

- The group mean (state here) is far away from the global (national) mean
- We have few observations for a unit (small N of counties here)

This shift is smaller when:

- The group mean is close to the global mean
- We have many observations for a unit (large N of counties)

# Summary

## What goes into a no-pooling (fixed effects) model

- Estimated intercepts are equal to empirical group means
- This approach typically *overfits* the data, and makes for poor prediction, unless we have a lot of data on each unit

## The partial pooling alternative

- The complete pooling approach assumes all units are the same
- The no-pooling approach assumes all units are different
- The *partial pooling* approach assumes that all units are different, but come from the same larger population

## Multilevel models are a good default

Multilevel models are very flexible and useful for modeling:

- Repeat observations of units over time
- Data with clusters

Multilevel models

- Shrink estimates toward group means
- Provide variance estimates across and within clusters
- Allow for flexible inference without overfitting

