---
title: "Manipulating and visualizing data in R"
author: "Frank Edwards"
institute: School of Criminal Justice, Rutgers - Newark
date: "2/1/2019"
output: binb::metropolis
---

```{r message = FALSE, warning = FALSE, echo=FALSE}
library(tidyverse)
def.chunk.hook  <- knitr::knit_hooks$get("chunk")
knitr::knit_hooks$set(chunk = function(x, options) {
  x <- def.chunk.hook(x, options)
  ifelse(options$size != "normalsize", paste0("\\", options$size,"\n\n", x, "\n\n \\normalsize"), x)
})
knitr::opts_chunk$set(warning=FALSE, message=FALSE, tidy = TRUE, size = "small")
cj_budgets<-read_csv("./data/lecture_2_demo.csv")
```

## Review HW1

- Challenges?

# Understanding the R environment

## Let's check installations

- R
- RStudio
- Git

## Let's clone the course repository

In RStudio Terminal, navigate to desired folder using command cd

https://github.com/f-edwards/intermediate_stats.git

## The R and RStudio environment

http://stat545.com/block002_hello-r-workspace-wd-project.html

## Using RMarkdown

http://stat545.com/block007_first-use-rmarkdown.html

## Data input/output 

http://stat545.com/block026_file-out-in.html

## Basics of data frames in R

http://stat545.com/block006_care-feeding-data.html

## Introducing dplyr

http://stat545.com/block009_dplyr-intro.html

## Model objects in R

- Work through basics of lm()
- Interpret output
- Extract objects from model object

# Working with (generalized) linear models

## Basic syntax of lm

To produce:

$$y = \beta_0 + \beta_1 X + \varepsilon$$
We use:

lm(y ~ x)

## Start a new R Project

1. Save it as lecture2.RProj in your intermediate_stats folder
2. move the data from ./data to ./lecture2/data (drag and drop or cp from terminal)

If using terminal (try it!) input

- mkdir data
- cd ..
- cp ./data/lecture_2_demo.csv ./lecture2/data/lecture_2_demo.csv


## Read in data

```{r}
cj_budgets<-read_csv("./data/lecture_2_demo.csv")
```

Note that we use a relative file path here (what's that!?)

## Explore the data

- Evaluate integrity with head()
- Evaluate column classes with str()
- Evaluate summary statistics with summary()

## Develop a theory for relationships between variables

Let's look the outcome \texttt{exp_police_pc}, which is the total expenditures on police per capita (divided by the total population) in a county across municipal and county governments (local PDs, sherrif's depts). 

**What could predict police budgets?**

```{r}
head(cj_budgets)
```

## Visualize the data, univariate

- hist()

## Visualize the data, bivariate

- Look at crosstabs with table()
- Using base, plot()

## Fit a single variable model

## Interpret the model

- What does each coefficient mean

## Visualize the fit

We can use coef() to extract coefficient estimates.

How could we use coef() to plot a fitted line?

## Fit a regression with multiple variables

- Interpret the coefficients
- Visualize the fit using coef() and plot()

## Fit a model with an interaction

- Interpret the coefficients
- Visualize the fit using coef() and plot()

## Introducing prediction and counterfactuals

- Create fake data
- Use coef() to generate expected values
- Use predict() to generate expected values

## Discuss HW2

- HW2 asks you to provide a brief analysis and writeup on the cj_budgets data
- Use RMarkdown to explore the data and write up your findings
- Construct a multivariate regression model and interpret your results
- Predict values for a theoretically interesting counterfactual
- Provide both criminological and statistical interpretations of your findings