---
title: "3: Basic Theory and Practice of Data Visualization"
author: "Frank Edwards"
date: "2/8/2019"
output: binb::metropolis
---

```{r echo=FALSE, message=FALSE}
library(tidyverse)
library(xtable)
options(xtable.comment = FALSE)
theme_set(theme_minimal())
def.chunk.hook  <- knitr::knit_hooks$get("chunk")
knitr::knit_hooks$set(chunk = function(x, options) {
  x <- def.chunk.hook(x, options)
  ifelse(options$size != "normalsize", paste0("\\", options$size,"\n\n", x, "\n\n \\normalsize"), x)
})
knitr::opts_chunk$set(warning=FALSE, message=FALSE, tidy = TRUE, size = "small")
```

## Data visualization

- What makes a good visual?
- Why visualize?
- How to use ggplot to make visuals in R

## Why do we visualize data?

```{r echo=FALSE}
### prep anscombe for small multiple plotting
### need a long dataframe with x,y,varname columns
data(anscombe)
plot_anscombe<-bind_cols(
  anscombe%>%
    select(x1,x2,x3,x4)%>%
    gather()%>%
    rename(x = value)%>%
    select(x),
  anscombe%>%
    select(y1,y2,y3,y4)%>%
    gather()%>%
    rename(y = value,
           var = key))
  
ggplot(plot_anscombe, 
       aes(x = x, y = y)) + 
  geom_smooth(method = "lm", 
              formula = y~ x) + 
  geom_point() + 
  facet_wrap(~var, scales = "free")

```

# Principles of good data visuals

## Good data visuals

- Are clearly labeled 
- Avoid deception
- Use repetition to invite comparisons
- Minimize 'chartjunk'

## Find the chartjunk
\begin{center}
		\includegraphics[width=\textwidth]{vis/excel_3d.pdf}
\end{center}

## The importance of axes
```{r echo = FALSE}
plot_dat<-data.frame("time" = 1:10,
                     "grade" = c(4, 3.9, 3.93, 3.9, 
                                 3.95,3.9, 3.88, 3.85, 
                                 3.8, 3.78))
ggplot(plot_dat, 
       aes(x = time, y = grade)) + 
  geom_line()
```

## The importance of axes 

```{r echo = FALSE}
ggplot(plot_dat, 
       aes(x = time, y = grade)) + 
  geom_line() + 
  coord_cartesian(ylim = c(3,4))
```

## The importance of aspect ratio

\begin{center}
  \includegraphics[width=\textwidth]{./vis/aspect_demo2.pdf}
\end{center}

# Why Visualize Data? 

## Why do we visualize data?

- Visuals can quickly reveal patterns in data
- Visuals are a (more) effective way to communicate quantitative information

# Geographic Data

##
```{r results = 'asis', echo = FALSE, size = "tiny"}
### load 1973 arrest rate data, built into R
data("USArrests")
### merge it onto map shape file
library(maps)
states<-map_data("state")
states<-left_join(states, 
                  USArrests%>%
                    mutate(region = 
                             tolower(
                               rownames(USArrests))))

tab_out<-as.data.frame(cbind(row.names(USArrests), USArrests$Murder))
xtable(tab_out)
```

##
```{r echo = FALSE}
ggplot(states, 
       aes(x = long, y = lat, 
           group = group, fill = Murder)) + 
  theme_minimal() + 
  geom_polygon()
```

# Which is most effective? Why?

# Time Series

## 
```{r echo = FALSE, results='asis', size = 'tiny'}
fe<-read_csv("./data/fe_1_25_19.csv")
fe_ts<-fe%>%
  group_by(`Date (Year)`)%>%
  summarise(deaths = n())%>%
  filter(`Date (Year)`<=2018)%>%
  mutate(`Date (Year)` = as.integer(`Date (Year)`))
xtable(fe_ts)
```

##
```{r echo = FALSE}
ggplot(fe_ts,
       aes(x = `Date (Year)`, y = deaths)) + 
  geom_line()
```

# Which is most effective? Why?

# Model results

## 

\begin{center}
		\includegraphics[width=\textwidth]{vis/ugly_table.png}
	\end{center}

## Reduced format: focal variable sign and significance 

\begin{table}[]
	\centering
	\tiny{
	\begin{tabular}{llllll}
	                                & Parameter        & All  & Violent & Drug  & Quality of life \\
	                                \hline
	Total                           & Between counties & +       & +           & +        & +                   \\
	                                & Within county    & +       & +           & +        & +                  \\
	African American                & Between counties & +       & +           & +        & +                   \\
	                                & Within county    & +       & +           & +        & +                   \\
	Asian-Pacific Islander          & Between counties & +       & +           & +        & +                   \\
	                                & Within county    & +       & +           &          & +                   \\
	American Indian / Alaska Native & Between counties & +       & +           & +        & +                   \\
	                                & Within county    & +       & +           & +        & +                   \\
	White                           & Between counties & +       & +           & +        & +                   \\
	                                & Within county    & +       & +           & +        & +                   \\
	                                \hline
	\end{tabular}
	}
	\end{table}

## Plot summary
\begin{center}
		\includegraphics[width=\textwidth]{vis/forest1.pdf}
	\end{center}
	
## Plot summary
\begin{center}
		\includegraphics[width=\textwidth]{vis/forest5.pdf}
	\end{center}
	
# Which is most effective? Why?

# Break

# Using ggplot2 to visualize data in R

## The importance of tidy (long) data for ggplot

Data is generally either wide or long

- In wide format, column position may indicate a variables value
- In long format, each variable has its own column

## Example of long data: each column is a variable

```{r}
head(iris)
```

## Example of the same data in wide format

```{r echo = FALSE, size = "tiny"}
species<-unique(iris$Species)
for(i in 1:length(species)){
  if(i==1){
    out<-iris%>%
      filter(Species==species[i])%>%
      select(-Species)
    names(out)<-paste(species[i],names(out), sep=".")
  }
  if(i>1){
    temp<-iris%>%
      filter(Species==species[i])%>%
      select(-Species)
    names(temp)<-paste(species[i],names(temp), sep=".")
    out<-bind_cols(out, temp)
  }
}

head(out)

```

## Tidy data lets us efficiently feed aesthetic parameters to ggplot.

- Tidy data is harder for humans to read in a spreadsheet, but much easier to program with. Tidyverse packages are built around making and keeping our R objects in tidy (long data.frame) format

- Try to keep your data tidy - all variables should be variables, not embedded in column names. 

**Frequent untidy variables:**

- Time (i.e. year)
- Group

# Basic anatomy of a ggplot command

## Give it data

```{r}
data("iris")
my_plot<-ggplot(data = iris)
```

## Add a single aesthetic parameter

```{r fig.height = 2, fig.width = 4, size = "scriptsize"}
ggplot(data = iris,
                aes(x = Petal.Width))
```

## Add a geom

```{r fig.height = 3, fig.width = 6, size = "scriptsize"}
ggplot(data = iris,
                aes(x = Petal.Width)) + 
  geom_histogram()
```

## Add two aesthetic parameters and a geom

```{r fig.height = 3, fig.width = 6, size = "scriptsize"}
ggplot(data = iris,
       aes(x = Petal.Width, 
           y = Petal.Length)) + 
  geom_point()
```

## Three variables: two continuous, one categorical 

```{r fig.height = 3, fig.width = 6, size = "scriptsize"}
ggplot(data = iris,
       aes(x = Petal.Width, 
           y = Petal.Length,
           color = Species)) +  
  geom_point()
```

## Three variables: two continuous, one categorical 

```{r fig.height = 3, fig.width = 6, size = "scriptsize"}
ggplot(data = iris,
       aes(x = Petal.Width, 
           y = Petal.Length,
           shape = Species)) +  
  geom_point()
```

## Multiple geoms

```{r fig.height = 3, fig.width = 6, size = "scriptsize"}
ggplot(data = iris,
       aes(x = Petal.Width, 
           y = Petal.Length,
           color = Species)) +  
  geom_point() + 
  geom_smooth(method = "lm")
```

## To review

ggplot needs three things to make a graphic

1. Data
2. Aesthetic paramaters
3. Geoms

# More advanced plots

## Boxplots (one continuous, one categorical)

```{r fig.height = 3, fig.width = 6, size = "scriptsize"}
ggplot(data = iris,
       aes(y = Petal.Width, 
           x = Species)) +  
  geom_boxplot()
```

## Violin plot 

```{r fig.height = 3, fig.width = 6, size = "scriptsize"}
ggplot(data = iris,
       aes(y = Petal.Width, 
           x = Species)) +  
  geom_violin()
```

## Small multiples (facets)

```{r fig.height = 3, fig.width = 6, size = "scriptsize"}
ggplot(data = iris,
       aes(x = Petal.Width, 
           y = Petal.Length)) +  
  geom_point() + 
  geom_smooth(method = "lm") + 
  facet_wrap(~ Species)
```

## Small multiples are very powerful

\begin{center}
  \includegraphics[width = \textwidth]{./vis/CI_ineq_map.pdf}
\end{center}

## Maps
```{r fig.height = 2, fig.width = 4, size = "tiny"}
data <- data.frame(murder = USArrests$Murder, 
                   state = tolower(rownames(USArrests)))
map <- map_data("state")
ggplot(data, aes(fill = murder)) + 
  geom_map(aes(map_id = state), map = map) +
  expand_limits(x = map$long, y = map$lat) 
```

