---
title: "3. Lab: Simulation"
author: "Frank Edwards"
institute: School of Criminal Justice, Rutgers - Newark
output: binb::metropolis
---

```{r message = FALSE, warning = FALSE, echo=FALSE}
library(tidyverse)
### configure for variable text size with chunk option
def.chunk.hook  <- knitr::knit_hooks$get("chunk")
knitr::knit_hooks$set(chunk = function(x, options) {
  x <- def.chunk.hook(x, options)
  ifelse(options$size != "normalsize", paste0("\\", options$size,"\n\n", x, "\n\n \\normalsize"), x)
})
knitr::opts_chunk$set(warning=FALSE, message=FALSE, tidy = TRUE, size = "small")
```

# Simulation in R

## Basic simulation functions: rbinom()

`rbinom()` takes random draws from a binomial distribution with arguments:

- n: number of observations
- size: number of trials per observation
- prob: probability of 'success' (1)

E.g. to flip one coin five times: 

```{r}
rbinom(n = 5, size = 1, prob = 0.5)
```

Or five coins five times:

```{r}
rbinom(n = 5, size = 5, prob = 0.5)
```

## Practice with Binomial simulation

Simulate recidivism after release. Assume that 10\% of subjects are re-arrested within 3 years of release. 
  - Simulate recidivism for 200 subjects. Compute the mean recidivism rate. 
  
```{r}
recid_sample<-rbinom(n= 200, size = 1, prob = 0.1)
recid_sample<-rbinom(200, 1, 0.1)
# how many ones
sum(recid_sample)
# compute the mean
sum(recid_sample)/200
mean(recid_sample)
```

Now let's assume that each of three years represents a trial, and we will count the number of years in which each subject had an arrest (size = 3).
  - Simulate recidivism for 200 subjects. Compute the mean number of years with an arrest
  
```{r}
recid_sample2<-rbinom(n= 200, size = 3, prob = 0.1)
### let's put it in a data frame
# wide first
recid_data<-data.frame(id = 1:200,
                       arrest_yr1 = rbinom(n= 200, size = 1, prob = 0.1),
                       arrest_yr2 = rbinom(n= 200, size = 1, prob = 0.1),
                       arrest_yr3 = rbinom(n= 200, size = 1, prob = 0.1))

### long
recid_data_long<-data.frame(id = rep(1:200, each = 3),
                            year = 1:3,
                            arrest = rbinom(n= 600, size = 1, prob = 0.1))

# was this person ever arrested
recid_data_summary<-recid_data_long %>% 
  group_by(id) %>% 
  summarize(ever_arrested = sum(arrest)>0)

```

  
## Practice with Normal simulation

Simulate income for US households. We will use a logarithmic scale because income inequality is generally multiplicative. 

Assume that income $x$ follows this distribution

$log(x) \sim N(\mu = 11, \sigma = 3)$

  - Simulate incomes for 100 households. 
  
```{r}
sample_households<-rnorm(n = 1000, mean = 6, sd = 3)
```

  - Compute a mean and standard deviation of your simulations

```{r}
mean(exp(sample_households))
sd(exp(sample_households))
```

  - Visualize your simulations using a histogram
  
```{r}
hist(exp(sample_households), breaks = 100)
```

  - Visualize your simulations on a dollar scale, rather than a log scale

## More complex simulation

Let's develop a simulation with a mixture of processes. 

- Create a data.frame with 500 rows, with `subject_n` as an integer 
- Assume a 0.25 probability of arrest, and simulate a new variable `arrest` in your data frame.
- Assume that log household income is systematically lower for individuals with an arrest as follows: $E[log(income_i)|arrest_i] = 11 - 2 \cdot arrest_i$
- Simulate household income, with $\mu$ as the expected income, and $\sigma = 3$
- Visualize

## Functions

Let's replicate our experiments using both loops and function mapping

Return to the recidivism experiment. 
- Write a custom function to sample recidivism for a single year (size = 1), allow n to be specified by the user

```{r eval = F}
# a demo function
waffles<-function(x){
  temp<-x + 2
  return(temp)
}
waffles(3)

### write a function that takes, then subtracts 3 and multiplies by 2

# evaluate 8, 3, -2
```

## Looping

- Write a for loop to draw 10 samples of 200 subjects. Store your output in a list

```{r eval = F}
# a demo loop
data_out<-list()
for(i in 1:50){
  data_out[[i]]<-waffles(i)
}


### run your custom function on all values between 100 and 200 and store output in a list

```

## Mapping

- Compute the mean for each of the samples

```{r eval = F}
# a demo of mapping a function to a list
map(data_out, mean)
```

## Mapping with a custom function

Return to the simple income simulation. 

- Write a custom function to generate samples of incomes 
- Use map to generate 20 samples of 50 household incomes
- Use map to compute a mean and standard deviation for each sample
- Use bind_rows to create a new data.frame with all of the observations. Be sure to include a sample index (i.e. 1-20)
- Visualize the distribution of incomes for each sample



## let's simulate bail decisions

```{r}
# let's use sample to simulate 200 bail decisions
decisions<-data.frame(
  decision = sample(c("bail", "no bail", "no bail sup", "jail"),
                  size = 200,
                  replace = T,
                  prob = c(0.19, 0.6, 0.2, 0.01)),
  id = 1:200)


```

