---
title: "Practice: Building a linear regression model"
author: "Frank Edwards"
institute: School of Criminal Justice, Rutgers - Newark
date: "1/23/2024"
output: binb::metropolis
---

```{r message = FALSE, warning = FALSE, echo=FALSE}
library(tidyverse)
def.chunk.hook  <- knitr::knit_hooks$get("chunk")
knitr::knit_hooks$set(chunk = function(x, options) {
  x <- def.chunk.hook(x, options)
  ifelse(options$size != "normalsize", paste0("\\", options$size,"\n\n", x, "\n\n \\normalsize"), x)
})
knitr::opts_chunk$set(warning=FALSE, message=FALSE, tidy = TRUE, size = "small")
```

## Model objects in R

- Work through basics of lm()
- Interpret output
- Simulate from model for inference

# Working with (generalized) linear models

## Basic syntax of lm

To produce:

$$y = \beta_0 + \beta_1 X + \varepsilon$$
We use:

lm(y ~ x)

# Are you in an R Project?

## Read in data

```{r}
cj_budgets<-read_csv("./data/lecture_3_demo.csv")
```

Note that we use a relative file path here (what's that!?)

## Explore the data

- Evaluate with head()
- Evaluate summary statistics with summary()

## Develop a theory for relationships between variables

Let's look the outcome exp_police_pc, which is the total expenditures on police per capita (divided by the total population) in a county across municipal and county governments (local PDs, sherrif's depts). 

**What could predict police budgets?**

## Visualize the data, univariate

- hist()

## Visualize the data, bivariate

- Look at crosstabs with table()
- Using base, plot()

## Fit a single variable model

## Interpret the model

- What does each coefficient mean

## Visualize the fit

We can use coef() to extract coefficient estimates.

How could we use coef() to plot a fitted line?

## Fit a regression with multiple variables

- Interpret the coefficients
- Visualize the fit using coef() and plot()

## Fit a model with an interaction

- Interpret the coefficients
- Visualize the fit using coef() and plot()

## Introducing prediction and counterfactuals

- Create fake data
- Use coef() to generate expected values
- Use predict() to generate expected values

## Discuss HW3

- HW2 asks you to provide a brief analysis and writeup on the cj_budgets data
- Use RMarkdown to explore the data and write up your findings
- Construct a multivariate regression model and interpret your results
- Predict values for a theoretically interesting counterfactual
- Provide both criminological and statistical interpretations of your findings