---
title: "Logistic regression, 1"
author: "Frank Edwards"
date: "2/19/2023"
output: binb::metropolis
---

```{r setup, echo = FALSE, message = FALSE}
library(tidyverse)
library(gridExtra)

theme_set(theme_bw())
def.chunk.hook  <- knitr::knit_hooks$get("chunk")
knitr::knit_hooks$set(chunk = function(x, options) {
  x <- def.chunk.hook(x, options)
  ifelse(options$size != "normalsize", paste0("\\", options$size,"\n\n", x, "\n\n \\normalsize"), x)
})
knitr::opts_chunk$set(warning=FALSE, message=FALSE, tidy = TRUE, size = "small")
```

## Read in the data for today

```{r size = "scriptsize"}
admissions <- read.csv("https://stats.idre.ucla.edu/stat/data/binary.csv")
head(admissions)
nrow(admissions)
```

## Evaluate distribution of binary admission variable

```{r}
table(admissions$admit)
mean(admissions$admit)
```

## Evaluate bivariate relationships between admission and focal predictors

```{r}
admissions %>% 
  group_by(admit) %>% 
  summarize(gpa_mean = mean(gpa),
            gre_mean = mean(gre))
```

## Lab: Let's fit a model

What else might predict admission?

```{r size = "tiny"}
head(admissions)
```

### 1. Theorize

Think carefully about what we want to accomplish. How do admissions work?

## 2. Write a model 

\[\textrm{admit} \sim \textrm{Bernoulli}(p)\]
\[\textrm{logit}(p_i) = \beta_0 + \beta_1\textrm{GPA}_i + \beta_2\textrm{GRE}_i\]

## 2. Fit the model

```{r}
### fit model
m1<-glm(admit ~ gpa + gre, 
             data = admissions,
             family = "binomial")

### create a counterfactual
gre <- c(200, 400, 600, 800)
gpa <- seq(from = 2, to = 4, by = 0.1)

fake_data <- expand_grid(gre, gpa)
### generate predictions
fake_data <- fake_data %>% 
  mutate(yhat = predict(m1,
                        newdata = fake_data,
                        type = "response"))
### visualize
p1<-ggplot(fake_data,
       aes(x = gpa, y = yhat,
           color = factor(gre))) + 
  geom_line()

#### THEN CRITIQUE AND REVISE THE MODEL

### they probably depend on each other
# write the math
# logit(pr(admit)) = b0 + b1 gpa + b2 gre + b3 gpa * gre

m2<-glm(admit ~ gpa * gre, 
             data = admissions,
             family = "binomial")

m3<-glm(admit ~ gpa + gre + factor(rank), 
             data = admissions,
             family = "binomial")

m4<-glm(admit ~ gpa * gre + factor(rank), 
             data = admissions,
             family = "binomial")


fake_data<-fake_data %>% 
  rename(yhat_m1 = yhat) %>% 
  mutate(yhat_m2 = predict(m2,
                           newdata = fake_data,
                           type = "response"))

ggplot(fake_data,
       aes(x = gpa, y = yhat_m1,
           color = factor(gre))) + 
  geom_line() + 
  labs(subtitle = "additive model")

ggplot(fake_data,
       aes(x = gpa, y = yhat_m2,
           color = factor(gre))) + 
  geom_line() + 
  labs(subtitle = "interactive model")


# ### define counterfactual scenarios
# potato2 <- data.frame(gpa = seq(from = 0, to = 4, by = 0.1),
#                       gre = 400) 
# 
# potato3 <- data.frame(gpa = seq(from = 0, to = 4, by = 0.1),
#                       gre = 600) 
# 
# cupcake <- bind_rows(potato2, potato3)
# ### estimate expected values using predict
# cupcake <- cupcake %>% 
#   mutate(yhat = predict(m1, 
#                         newdata = cupcake, 
#                         type = "response"))
# ### visualize
# ggplot(cupcake,
#        aes(x = gpa, y = yhat, color = factor(gre))) + 
#   geom_line()


### same thing
# yhat <- predict(m1, newdata = potato2,
#         type = "response")



```

## 3. Compare model fits

```{r}
## baseline gpa model
m0<-glm(admit ~ gpa,
        data = admissions,
        family = "binomial")




```

## 3. Comparing model fits

```{r}
AIC(m1, m0)
```

## Iterate

Repeat steps 1-3 until satisfied that we've built a theoretically justified well-fitting model. 

## 4. Interpet

1. Identify what we'd like to make inferences about. Which groups or cases would be theoretically interesting for our topic.

2. Create these cases in a new data frame.

3. Predict

4. Visualize

## Homework 5

Who was most (and least) likely to die on the Titanic? Use `./data/titanic.csv` for this one.

1. Develop a theory for survival
2. Write a model
3. Fit the model
4. Think about revising the model
5. Compare model fits
6. Interpret the model 

```{r}
dat<-read_csv("./data/titanic.csv")

m1<-glm(Survived ~ Fare,
        family = "binomial",
        data = dat)


```




```{r}
dat<-read_tsv("cause_of_death.txt")

dat<-dat %>% 
  mutate(Deaths = as.numeric(Deaths)) %>% 
  select(State, Gender, 
         `Single Race 6`, Deaths, 
         Population) %>% 
  rename(race = `Single Race 6`) %>% 
  mutate(crude_rate = Deaths / Population * 1e5)

dat_state<-dat %>% 
  group_by(State) %>% 
  summarize(Deaths = sum(Deaths),
            Population = sum(Population)) %>% 
  mutate(crude_rate = Deaths / Population)

ggplot(dat_state, aes(x = crude_rate)) + 
  geom_histogram() 
```

