---
title: "Working with categorical variables"
author: "Frank Edwards"
output: binb::metropolis
---

```{r setup, echo = FALSE, message = FALSE}
library(tidyverse)
library(broom)
set.seed(1)

options(xtable.comment = FALSE)
theme_set(theme_minimal())
def.chunk.hook  <- knitr::knit_hooks$get("chunk")
knitr::knit_hooks$set(chunk = function(x, options) {
  x <- def.chunk.hook(x, options)
  ifelse(options$size != "normalsize", paste0("\\", options$size,"\n\n", x, "\n\n \\normalsize"), x)
})
knitr::opts_chunk$set(warning=FALSE, message=FALSE, tidy = TRUE, size = "small")
```

# Review HW 5

# Working with strings in R

## What is a string?

*A string is an ordered sequence of characters*

- Strings are generally stored verbatim, and have no mathematical meaning (ie math operations will return errors)
- In R, these are character objects
- Generally wrapped in ""
- In R, can use as.character to convert any value to character

## Let's try something

What's the difference between these commands

```{r}
a<-1
b<-"1"
```

- What does each command do?
- Try str() on each

## Let's keep trying to break R

On your console, try these:

- a<-c(1,2,3)
- b<-c("1", "2", "3")
- c<-c(1, "2", 3)
- d<-c(1, 2, "c") \pause

What happened? Why?

## Let's keep trying to break R

- a + a
- a + "a"
- a + b  
- "a" == "A"
- a == "a"

## Summary of strings in R

- R will coerce vectors to string when strings are included
- Strings are the most complex variable type (in order: logical, numeric, factor, character)
- Strings can only be compared to strings
- You should generally treat all categorical variables as strings in R (unless order matters! then use factor())

# Working with strings in R

## Stringr

The stringr package loads with tidyverse

```{r}
library(tidyverse)
```

It has more powerful versions of base functions like:

- substr()
- grep()
- paste()
- strsplit()

## Getting started

```{r}
word<-"banana"
str_length(word)
word%>%
  str_length()
```

## Pulling single characters from a string

```{r}
word<-"banana"
word_length<-str_length(word)
word%>%
  str_sub(1, 1)
```
\pause
```{r}
for(i in 1:word_length){
  print(str_sub(word, i, i))
}
```

## Pulling multiple characters
```{r}
word<-"banana"
word%>%
  str_sub(1, 3)
```
\pause
```{r}
for(i in 1:word_length){
  print(str_sub(word, 1, i))
}
```

## Substitution
```{r}
str_sub(word, 1, 2)<-"surprise"
word
```

## Indexing on strings, negative values

```{r}
word
str_sub(word, -2, -1)
```

What happened here?

## Some convenient functions

```{r}
phrase<-"bananas are the tastiest"
toupper(phrase)
```
\pause
```{r}
tolower(toupper(phrase))
```
\pause
```{r}
library(tools)
toTitleCase(phrase)
```
\pause
```{r}
odd<-"   bananas are the tastiest     "
trimws(odd)
```

## Splitting a string

```{r, size = "tiny"}
str_split(phrase, pattern = " ")
```
\pause
```{r, size = "tiny"}
str_split(phrase, pattern = "a")
```

## Splitting a string to a fixed matrix
```{r, size = "tiny"}
str_split_fixed(phrase, pattern = " ", n = 2)
```
\pause

```{r, size = "tiny"}
str_split_fixed(phrase, pattern = " ", n = 3)
```
\pause
```{r, size = "tiny"}
str_split_fixed(phrase, pattern = " ", n = 4)
```

## Finding strings in strings

```{r}
phrase
str_detect(phrase, "are")
```

\pause

```{r}
str_detect(phrase, "scrumptious")
str_detect(phrase, "nana")
```

## Squishing strings together

```{r}
str_c(phrase, "seriously")
### oops
str_c(phrase, "seriously", sep = " ")
### or 
str_c(phrase, " seriously")
## not this
str_c(phrase, "seriously", sep = "!!")
```


## But we usually work with vectors!

- This is true

- All of this works on vectors 

- Like a vector of fruits! \pause

```{r, size = "tiny"}
fruit
```

## See, it works on vectors!

```{r, size = "tiny"}
str_sub(fruit, 1, 2)
```

## Let's see how many fruits use the word "fruit"

```{r size = "tiny"}
fruit%>%
  str_detect("fruit")
```
\pause
```{r size = "tiny"}
## How many?
fruit%>%
  str_detect("fruit")%>%
  sum()
```

## Let's get those fruits that are called fruits

```{r}
fruitfruits<-str_subset(fruit, "fruit")
```

## Let's make them all one word

```{r, size = "scriptsize"}
fruitfruits<-str_replace(
  
  fruitfruits, 
  pattern = " ", 
  replacement= "")

fruitfruits
```

## Let's make them all two words

```{r size = "scriptsize"}
fruitfruits<-str_replace(
  
  fruitfruits, 
  pattern = "fruit", 
  replacement = " fruit")

fruitfruits
```

## Using str_replace to handle NA
```{r size = "scriptsize"}
melons <- str_subset(fruit, pattern = "melon")
melons[2] <- NA
melons
#> [1] "canary melon" NA             "watermelon"
str_replace_na(melons, "UNKNOWN MELON")
```

# Moving on to some more practical examples

## Returning to the Titanic

```{r}
titanic<-read_csv("./data/titanic.csv")
```

## A handy trick
```{r}
tolower(names(titanic))
names(titanic)<-tolower(names(titanic))
```

## Let's see what titles people used

```{r, size = "tiny"}
titanic_titles<-titanic%>%
  separate(name, 
           into = c("title", "name"),
           sep = "\\.")
## the \\ is there because . has a special meaning in regex (we'll come back to that)
titanic
```

## Titles on the Titanic

```{r size = "scriptsize"}
unique(titanic_titles$title)
```

## Who's Jonkheer? Who's the Countess?

```{r}
grep("Jonkheer", titanic$name)
grep("the Countess", titanic$name)
```

## grep and grepl

Both use regular expressions to match patterns in strings. 

- grep() returns the index of matches (ie row number)
- grepl() returns TRUE or FALSE for matches
- Regular expressions (or regex) are super powerful and super confusing. 
- Here's a cheat sheet (https://www.rstudio.com/wp-content/uploads/2016/09/RegExCheatsheet.pdf)
- Most of the time, we don't need to worry about regex. But special characters can trip you up.

*Special characters in regex*

    \^$.|?*+()[{"  

## grep and grepl in practice

How many countesses are there?

```{r}
table(grepl("the Countess", titanic$name))
```

## Who was she?!

```{r, size = "scriptsize"}
titanic[grep("the Countess", titanic$name),]

titanic%>%
  filter(grepl("the Countess", titanic$name))
```

## Let's just get her title and name

```{r, size = "scriptsize"}
titanic%>%
  filter(grepl("the Countess", titanic$name))%>%
  select(name)
```

# Recoding - ifelse and case_when

## The ifelse() function

ifelse() commands require the following:

1. test: a conditional statement that returns TRUE or FALSE
2. yes: a value assigned when test==TRUE
3. no: a value assigned when test==FALSE

## if and else

```{r, size = "scriptsize"}
a<-c(1,2)
b<-c(1,2)
if(a==b){
  "equal!"
} else{
  "not equal!"
}

if(a!=b){
  "not equal!"
} else{
  "equal!"
}
```

## What if we want a comparison of each element in the vector?

```{r}
ifelse(a==b, "equal!", "not equal!")
```

We can use this to do all kinds of neat things.

## We're going to be cruel for a moment

Let's add "You died" to the front of any the name of any passenger who died

```{r}
cruelty<-titanic%>%
  mutate(
    
    name = 
      ifelse(
        survived == 0,
        str_c("You died", name, sep = " "),
        name)
  )%>%
  select(survived, name)
```

## What did it do?

```{r size = "tiny"}
cruelty
```

## Let's add a new variable - child

```{r size = "scriptsize"}
kids<-titanic%>%
  mutate(
    
    child = ifelse(age<18,
                   "Child",
                   "Adult"))

table(kids$child)
table(titanic$age<18)
```

## Let's recode the variable sex

```{r}
recode<-titanic%>%
  mutate(
    
    sex = ifelse(sex == "male",
                      "m",
                      "f"))
```

## But what if we have more than one condition to evaluate?

Let's make a three category age variable: child, adult, elder

We could nest ifelse() commands:

```{r}
age_recode<-titanic%>%
  mutate(age_cat = 
           
           ifelse(age<18, "child",
                  ifelse(age>65,
                         "elder",
                         "adult")))

table(age_recode$age_cat)
```

## But that's hard to read and can get cumbersome with many categories

case_when() is a flexible approach to link together many conditional statements

```{r, size = "scriptsize"}
age_recode2<-titanic%>%
  mutate(age_cat = 
           
           case_when(
             age < 18 ~ "child",
             age >= 18 & age <= 65 ~ "adult",
             age >65 ~ "elder"
               
           ))

table(age_recode2$age_cat)
```

## A real example: HW 6

```{r size = "tiny"}
fe<-read_csv("./data/fe_1_25_19.csv")
unique(fe$`Subject's age`)
```

## More messy data

```{r size = "scriptsize"}
unique(fe$`Subject's gender`)
```

## Data cleaning

- Data cleaning involves writing code to solve problems in the raw data. 
- We write programs that search out and fix issues so that we can conduct needed analysis.
- NEVER modify the original data. Doing so is not reproducible or documented.

