---
title: "HW1"
subtitle: "Intermediate Statistics"
author: "Frank Edwards"
date: "1/27/2021"
output: html_document
---

```{r include = F}
### set to project root directory rather than rmd folder
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())

### load packages
library(tidyverse)
```

## Question 1

For this problem, we will use data built into R. Load the data using the console command data(anscombe). 

```{r}
data("anscombe")
```


- Describe the relationships between y1 and x1, y2 and x2, y3 and x3, and y4 and x4 in sequence using a correlation coefficient. 

```{r}
with(anscombe, cor(x1, y1))
with(anscombe, cor(x2, y2))
with(anscombe, cor(x3, y3))
with(anscombe, cor(x4, y4))

### instead of with() could also use anscombe$x1, anscombe$y1
```



- Construct linear regression models for each pair. Describe your results.

```{r}
### estimate models
m1<-lm(y1~x1, data = anscombe)
m2<-lm(y2~x2, data = anscombe)
m3<-lm(y3~x3, data = anscombe)
m4<-lm(y4~x4, data = anscombe)
### show model coefficient estimates
coef(m1)
coef(m2)
coef(m3)
coef(m4)

### whoa they are all the same...
```


- Draw a scatterplot for each pair

```{r}
ans_df<-data.frame(anscombe)
ggplot(ans_df, aes(x = x1, y = y1)) + 
  geom_point()

ggplot(ans_df, aes(x = x2, y = y2)) + 
  geom_point()

ggplot(ans_df, aes(x = x3, y = y3)) + 
  geom_point()

ggplot(ans_df, aes(x = x4, y = y4)) + 
  geom_point()
```


- Summarize your findings

Weird! Always look at your data! Don't rely on summary statistics!

## Question 2

Use the data in revenue_dat.csv in the course repository. 

1. Load in the data

```{r include = F}
cj_budgets<-read_csv("./hw/data/revenue_dat.csv")
```

2. Use the mutate() function in the dplyr package to construct a new variable for property tax revenue per capita called prop_tax_pc using the variables rev_prop_tax and pop_tot

```{r}
cj_budgets<-cj_budgets %>% 
  mutate(prop_tax_pc = rev_prop_tax / pop_tot)
```

3. Transform prop_tax_pc into a binary variable equal to TRUE if property tax revenues per capita are greater than or equal to the median value of the variable, FALSE otherwise

```{r}
cj_budgets<-cj_budgets %>% 
  mutate(prop_tax_pc = prop_tax_pc>=median(prop_tax_pc))
```


4. Run a linear regression model of the proportion of the population in deep poverty (pop_pct_deep_pov) as a function of property tax revenues per capita (your binary measure).

```{r}
m1<-lm(pop_pct_deep_pov ~ prop_tax_pc, 
       data = cj_budgets)
```


## Question 3

Interpret your findings from the regression model.

Counties with below median property tax revenue in the data had a deep poverty rate of about `r round(coef(m1)[1],3)*100` (equal to $\beta_0 \times 100$) percent of the population. Counties with above median property tax revenue in the data had a deep poverty rate of about `r round(coef(m1)[1] + coef(m1)[2],3) * 100` percent of the population ($\beta_0 + \beta_1 \times 100$). The difference between high and low property tax revenue counties is equal to $\beta_1$.