---
title: "HW1"
subtitle: "Intermediate Statistics"
author: "Frank Edwards"
date: "1/27/2021"
output: html_document
---

```{r include = F}
### set to project root directory rather than rmd folder
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())

### load packages
library(tidyverse)
```

## Question 1

For this problem, we will use data built into R. Load the data using the console command data(anscombe). 

- Describe the relationships between y1 and x1, y2 and x2, y3 and x3, and y4 and x4 in sequence using a correlation coefficient. 
- Construct linear regression models for each pair. Describe your results.
- Draw a scatterplot for each pair
- Summarize your findings

## Question 2

Use the data in revenue_dat.csv in the course repository. 

1. Load in the data

```{r include = F}
cj_budgets<-read_csv("./hw/data/revenue_dat.csv")
```


2. Use the mutate() function in the dplyr package to construct a new variable for property tax revenue per capita called prop_tax_pc using the variables rev_prop_tax and tot_pop.

3. Transform prop_tax_pc into a binary variable equal to TRUE if property tax revenues per capita are greater than or equal to the median value of the variable, FALSE otherwise

4. Run a linear regression model of the proportion of the population in deep poverty (pop_pct_deep_pov) as a function of property tax revenues per capita (your binary measure).

## Question 3

Interpret your findings from the regression model.
