---
title: "HW1"
subtitle: "Intermediate Statistics"
author: "YOU"
date: "SUBMISSION DATE"
output: html_document
---

## Question 1

For this problem, we will use data built into R. Load the data using the console command data("anscombe"). 

- Describe the relationships between y1 and x1, y2 and x2, y3 and x3, and y4 and x4 in sequence using a correlation coefficient. 

```{r}
library(tidyverse)
data(anscombe)

anscombe %>% 
  summarize(cor(y1, x1),
            cor(y2, x2),
            cor(y3, x3),
            cor(y4, x4))
```

They are the same!

- Construct linear regression models for each pair. Describe your results.

```{r}
m1<-lm(y1~x1, data = anscombe)
m2<-lm(y2~x2, data = anscombe)
m3<-lm(y3~x3, data = anscombe)
# etc
```

they are the same!

- Draw a scatterplot for each pair

```{r}
data(anscombe)

ggplot(anscombe,
       aes(x = x1,
           y = y1)) + 
  geom_point()

ggplot(anscombe,
       aes(x = x2,
           y = y2)) + 
  geom_point()

ggplot(anscombe,
       aes(x = x3,
           y = y3)) + 
  geom_point()

ggplot(anscombe,
       aes(x = x4,
           y = y4)) + 
  geom_point()

# etc
```

they are not the same!

- Summarize your findings

We should visualize our data!

## Question 2

Use the data in revenue_dat.csv in the course repository.  

1. Load in the data with read_csv(). 

```{r}
dat<-read_csv("https://github.com/f-edwards/intermediate_stats/raw/master/data/revenue_dat.csv")
```

2. Create a new data frame that subsets the data using select() to only the following columns: fips_st, fips_cnty, pop_tot, rev_prop_tax, pop_pct_deep_pov

```{r}
dat<-dat %>% 
  select(fips_st, 
         fips_cnty, 
         pop_tot, 
         rev_prop_tax, 
         pop_pct_deep_pov)
```

3. Use the mutate() function in the dplyr package to construct a new variable for property tax revenue per capita called prop_tax_pc using the variables rev_prop_tax and pop_tot

```{r}
dat<-dat %>% 
  mutate(prop_tax_pc = rev_prop_tax / pop_tot)
```

4. Run a linear regression model with the proportion of the population in deep poverty (pop_pct_deep_pov) as an outcome and property tax revenues per capita (prop_tax_pc) as a predictor. 

```{r}
library(broom)
m2<-lm(pop_pct_deep_pov ~ log(prop_tax_pc), data = dat)
```

5. Interpret your findings from the regression model. Explain the meaning of the coefficients as well as the real-world meaning of the relationship you've modeled. Length should be 1-2 paragraphs.

There's a negative relationship between property taxes and poverty. For each $1 increase in prop tax pc, there's small decrease in expected deep poverty.

This means counties with larger property tax bases have much lower deep poverty rates. Inequality!