---
title: "HW2"
author: "Frank Edwards"
date: "1/31/2020"
output: html_document
---

**Due via email by 10AM 2/7 to cs1215@scj.rutgers.edu**

Please use RMarkdown to complete this homework, and knit it into an html report. You can write your code and responses inline. For help with RMarkdown, see [this guide](https://rmarkdown.rstudio.com/lesson-1.html), or the cheat sheet in RStudio's menus: Help->cheatsheets->R Markdown Cheat Sheet.

The easy questions are worth a total of 6 points. Hard questions are worth a total of 4 points.

*Note:* You may use either base R code (as in the book) or tidyverse code (as in my slides and lab).

## Easy

These problems use the samples from the posterior distribution for the globe tossing example.
This code will give you a specific set of samples, so that you can check your answers exactly.

```{r}
p_grid <- seq( from=0 , to=1 , length.out=1000 )
prior <- rep( 1 , 1000 )
likelihood <- dbinom( 6 , size=9 , prob=p_grid )
posterior <- likelihood * prior
posterior <- posterior / sum(posterior)
# set.seed() sets the random seed your computer uses for sampling, guaranteeing the same results
set.seed(100) 
samples <- sample( p_grid , prob=posterior , size=1e4 , replace=TRUE )
```

Use the values in samples to answer the questions that follow.

**3E1.** How much posterior probability lies below p = 0.2?

**3E2.** How much posterior probability lies above p = 0.8?

**3E3.** How much posterior probability lies between p = 0.2 and p = 0.8?

**3E4.** 20% of the posterior probability lies below which value of p?

**3E5.** 20% of the posterior probability lies above which value of p?

**3E6.** Which values of p contain 66% of the posterior probability, assuming equal posterior probability both below and above the interval?

## Hard

**Introduction.** The practice problems here all use the data below. These data indicate the gender (male=1, female=0) of officially reported first and second born children in 100 two-child families.

```{r message = F}
library(rethinking)
data(homeworkch3)
birth1
birth2
```

So for example, the first family in the data reported a boy (1) and then a girl (0). The second family reported a girl (0) and then a boy (1).

Use these vectors as data. So for example to compute the total number of boys born across all of these births, you could use:

```{r}
sum(birth1) + sum(birth2)
```

**H1:** Using grid approximation, compute the posterior distribution for the probability of a birth being a boy. Assume a uniform prior probability. Which parameter value maximizes the posterior probability?

```{r}
p_grid <- seq( from=0 , to=1 , length.out=1000 )
prior <- rep( 1 , 1000 )
likelihood <- dbinom(sum(birth1) + sum(birth2) , size=length(birth1) + length(birth2) , prob=p_grid )
posterior <- likelihood * prior
posterior <- posterior / sum(posterior)
# set.seed() sets the random seed your computer uses for sampling, guaranteeing the same results
set.seed(100) 
samples <- sample( p_grid , prob=posterior , size=1e4 , replace=TRUE )
```

**H2:** Using the sample function, draw 10,000 random parameter values from the posterior distribution you calculated above. Use these samples to estimate the 50%, 89%, and 97% central percentile intervals.

**H3:** Use rbinom to simulate 10,000 replicates of 200 births. You should end up with 10,000 numbers, each one a count of boys out of 200 births. Compare the distribution of predicted numbers of boys to the actual count in the data (111 boys out of 200 births). There are many good ways to visualize the simulations, but the dens command (part of the rethinking package) is probably the easiest way in this case. Does it look like the model fits the data well? That is, does the distribution of predictions include the actual observation as a central, likely outcome?

```{r}
sims<-rbinom(1e4, 200, prob = samples)
```

**H4.** Now compare 10,000 counts of boys from 100 simulated first borns only to the number of boys in the first births, birth1 . How does the model look in this light?


**H5.** The model assumes that sex of first and second births are independent. To check this assumption, focus now on second births that followed female first borns. Compare 10,000 simulated counts of boys to only those second births that followed girls. To do this correctly, you need to count the number of first borns who were girls and simulate that many births, 10,000 times. Compare the counts of boys in your simulations to the actual observed count of boys following girls. How does the model look in this light? Any guesses what is going on in these data?

```{r}
sims<-rbinom()
```

