---
title: "HW3"
author: "Frank Edwards"
date: "2/7/2020"
output: html_document
---

**Due via email by 10AM 2/14 to cs1215@scj.rutgers.edu**

```{r setup, include=FALSE}
library(tidyverse)
library(rethinking)
```

The easy questions are worth a total of six points. The hard questions are worth a total of 4 points. You can earn a total of 10 points on this homework.

# Easy

**E1:** In the model definition below, which line is the likelihood?

1. $y_i \sim Normal(\mu, \sigma)$
2. $\mu \sim Normal(0, 10)$
3. $\sigma \sim Uniform(0,10)$

**E2:** In the model definition above, how many parameters are in the posterior distribution?

**E3:** In the model definition below, which line is the linear model?

1. $y_i \sim Normal(\mu, \sigma)$
2. $\mu_i = \alpha + \beta x_i$
3. $\alpha \sim Normal(0,10)$
3. $\beta \sim Normal(0,10)$
3. $\sigma \sim Uniform(0,10)$

**E4:** In the model definition above, how many parameters are in the posterior distribution?

# Hard

These questions use the mtcars data built into R 

```{r}
data(mtcars)
head(mtcars)
```

**H1:** A car's fuel efficiency (mpg) is probably related to it's weight (wt). Using quap, estimate a linear regression with mpg as the outcome, and weight as a predictor. Use the following model specification, where $y$ is mpg and $x$ is weight:

\[y_i \sim Normal(\mu, \sigma)\]
\[\mu = \alpha + \beta x_i\]
\[\alpha \sim Normal(0,10)\]
\[\beta \sim Normal(0,10)\]
\[\sigma \sim Uniform(0,10)\]

Describe the posterior distribution (hint: summary() provides basic information on a quap object, extract.samples() provides draws of each parameter from the posterior).

```{r}
h1<-quap(
  flist = alist(
    mpg ~ dnorm(mu, sigma),
    mu<-a + b * wt,
    a ~ dnorm(0,10),
    b ~ dnorm(0,10),
    sigma ~ dunif(0,10)
  ),
  data = mtcars
)

summary(h1)

h1_samples<-extract.samples(h1, n=2)
dens(h1_samples$a)
dens(h1_samples$b)
dens(h1_samples$sigma)
```

**H2:** Use prior predictive simulation to generate predicted fuel efficiencies for 10,000 cars with a weight of 3 tons (hold weight constant at 3) from the prior. Visualize the distribution of these predicted car fuel efficiencies with a density plot or histogram. Are these sensible priors?

```{r}
a<-rnorm(1e4, 0, 10)
b<-rnorm(1e4, 0, 10)
sigma<-runif(1e4, 0, 10)
weight<-3
mpg_prior_preds<-rnorm(1e4, mean = a + b * weight, sd = sigma)
```

**H3:** Revise your priors such that prior predictions have reasonable bounds. Explain your choices.

```{r}
a<-rnorm(1e4, 25, 7)
b<-rnorm(1e4, -3, 2)
sigma<-runif(1e4, 0, 10)
weight<-3
mpg_prior_preds<-rnorm(1e4, mean = a + b * weight, sd = sigma)
hist(mpg_prior_preds)
```

**H4:** Estimate the regression model in H1 with these new priors. Compare the posterior distributions from H1 and H5. What differences do you see?

```{r}
h5<-quap(
  flist = alist(
    mpg ~ dnorm(mu, sigma),
    mu<-a + b * wt,
    a ~ dnorm(25, 7),
    b ~ dnorm(-3,2),
    sigma ~ dunif(0,10)
  ),
  data = mtcars
)

summary(h5)

h5_samples<-extract.samples(h5)
dens(h5_samples$a)
dens(h5_samples$b)
dens(h5_samples$sigma)
```