---
title: "Lab - R as a programming language"
author: "Frank Edwards"
date: "4/14/2020"
output: ioslides_presentation
---

## Loops

```{r}
for(i in 1:10){
  print("Hello world")
  print(i)
}
```

## Nested loops

```{r}
for(j in 1:2){
  for(i in 1:3){
    print(paste("i=", i, ", j=", j))
  }
}
```

## Loops as vector operations

- Set up a vector named x with length 10, where each value is a random numeric variable
- Use a loop (and matrix indexing) to create a new vector called y, where each value $y_i$ is equal to $2x_i$
- Then tell me a *much* easier way to do this

## File reading demo

- Open or source ./lab/lab_programming_setup.r 
- Write a script to read in each file
- *Hint:* list.files("path") creates a list of files in a given directory

## Functions

Let's make a vector of 10 random integers between 1 and 10

```{r}
x<-rbinom(n = 10, size = 10, prob =  0.5)
x
```

## Functions

Let's add 1 to each value, then divide by 2

```{r}
my_first_function<-function(variable){
  temp<-variable + 1
  temp<- temp / 2
  return(temp)
}

x
my_first_function(x)
```

## More advanced mathematical functions

The Normal PDF is 

\[\frac{1}{\sigma\sqrt{2\pi}} e^{-\frac{1}{2}\left(\frac{x - \mu}{\sigma}\right)^2}\]

Write a function to compute the probability density of any number $x$ for a given value of $\mu$ and $\sigma$

## Functions for data cleaning

Let's say that each teacher in our example data was someone from Star Wars, but we've temporarily called labeled them with integers.

- load data(starwars) from tidyverse
- The name is the name of each of our teachers, in order
- Write a function to assign each integer value in the test score data to the appropriate teacher
